---
title: "Getting started with nanowasm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with nanowasm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

nanowasm runs [WebAssembly](https://webassembly.org/) modules from R. The
interpreter, [toywasm](https://github.com/yamt/toywasm), is bundled with the
package, so there is nothing else to install.

This vignette walks through the whole API with the small example modules
installed with the package: loading a module, calling its functions, moving
data through its memory, giving it R functions to call, and limiting what it
can consume.

```{r setup}
library(nanowasm)
example <- function(name) system.file("extdata", name, package = "nanowasm")
```

## Modules and instances

A *module* is compiled WebAssembly code: a `.wasm` file, or the same bytes in
a raw vector. `wasm_module()` decodes and validates it, and printing it shows
what it exports and imports:

```{r}
mod <- wasm_module(example("fib.wasm"))
mod
```

An *instance* is a module brought to life, with its own memory, tables and
globals. Calls go to an instance:

```{r}
inst <- wasm_instantiate(mod)
inst$fib(20L)
wasm_call(inst, "fib", 20L)
```

A module can be instantiated as many times as you like, and instances don't
share state.

## Values

WebAssembly functions take and return four number types. They map to R like
this:

| WebAssembly | from R | to R |
|-------------|--------|------|
| `i32` | integer, or a whole double in \[-2^31^, 2^32^) | integer |
| `i64` | a whole number with magnitude at most 2^53^ | double |
| `f32`, `f64` | integer or double | double |

Arguments are checked against the signature, and a mismatch is an error
rather than a silent coercion:

```{r, error = TRUE}
add <- wasm_instantiate(wasm_module(example("add.wasm")))
add$add(2L, 3L)
add$add_f64(0.5, 0.25)
add$add(2.5, 1L)
```

A function with several results returns them as a list, and one with none
returns `NULL` invisibly.

## Linear memory

Calls pass numbers only. Anything larger, such as a vector or a string,
travels through the instance's *linear memory*, a byte array that the
module reads and writes with pointers (byte offsets). `wasm_memory()` gives
R the same view:

```{r}
inst <- wasm_instantiate(wasm_module(example("sum.wasm")))
mem <- wasm_memory(inst)
mem
```

WebAssembly has no standard allocator, so R has to ask the module where it
may write. `sum.wasm` exports a simple `alloc()`:

```{r}
x <- c(1.5, 2.5, 3, 4, 5)
ptr <- inst$alloc(8L * length(x))
wasm_write(mem, ptr, x, type = "f64")
inst$sum_f64(ptr, length(x))
wasm_read(mem, ptr, length(x), type = "f64")
```

`type` can be `"raw"`, `"i8"`, `"u8"`, `"i16"`, `"u16"`, `"i32"`, `"u32"`,
`"i64"`, `"u64"`, `"f32"` or `"f64"`. Offsets are 0-based byte addresses, the
numbers the module itself uses, not 1-based R indices. Every access is
bounds-checked:

```{r, error = TRUE}
wasm_read(mem, wasm_memory_size(mem) - 2, 4)
```

Strings are UTF-8:

```{r}
end <- wasm_write_string(mem, 2048, "héllo")
wasm_read_string(mem, 2048)
```

Exported globals are read with `wasm_global()` and set with
`wasm_global<-`:

```{r}
wasm_global(inst, "heap_top")
```

## Calling R from WebAssembly

A module can declare *imports*: functions it expects its host to provide.
`wasm_imports()` lists them:

```{r}
logmod <- wasm_module(example("log.wasm"))
wasm_imports(logmod)
```

Provide them as R functions wrapped with `wasm_func()`, which records the
signature, in a list of lists named by import module and field. A function
with a `caller` argument can reach the calling instance's memory, which is
how this one reads the string the module passes by pointer and length:

```{r}
log <- wasm_func(
  function(ptr, len, caller) {
    message("module says: ", wasm_read_string(caller$memory(), ptr, len))
  },
  params = c("i32", "i32")
)
inst <- wasm_instantiate(logmod, imports = list(env = list(log = log)))
inst$greet()
```

Linking checks every import before anything runs, and reports all missing
or mismatched imports together:

```{r, error = TRUE}
wasm_instantiate(logmod, imports = list(env = list(log = wasm_func(identity, "i32"))))
```

If the R function fails, the WebAssembly call stops with a
`nanowasm_host_error` that carries the original condition as `parent`.
Interrupts, restarts and other jumps out of the R function also unwind the
WebAssembly call safely, and the instance stays usable.

## Running WASI programs

Programs compiled for WASI, the WebAssembly System Interface (C with
[wasi-sdk](https://github.com/WebAssembly/wasi-sdk), Rust's `wasm32-wasip1`
target, and others), expect a small operating system: arguments,
environment variables, standard input and output, clocks, random numbers and
files. `wasm_run()` provides one, implemented in R, and runs the program:

```{r}
hello <- example("hello-wasi.wasm")
res <- wasm_run(hello, args = c("from", "R"), env = c(GREETING = "Hi"))
res
```

A program sees only what you give it. Files are reachable only inside the
directories you grant with `dirs`, read-only unless `writable = TRUE`:

```{r}
dir <- tempfile()
dir.create(dir)
writeLines(c("x,y", "1,2"), file.path(dir, "data.csv"))

cat_wasm <- example("cat-wasi.wasm")
wasm_run(cat_wasm, args = "data.csv", dirs = dir)$stdout
wasm_run(cat_wasm, args = "../anything.txt", dirs = dir)
```

For more control, build the environment with `wasm_wasi()`, instantiate the
module with it, and call `wasm_wasi_start()`, or call the exports of a
WASI "reactor" module directly. Random numbers come from R's generator, so
`set.seed()` makes a program reproducible.

## Limits and errors

Every instance runs under limits set by `wasm_limits()`: memory, call depth,
value-stack size and time per call.

```{r, error = TRUE}
slow <- wasm_instantiate(
  wasm_module(example("fib.wasm")),
  limits = wasm_limits(memory = 16 * 2^20, timeout = 0.5)
)
slow$fib(40L)
```

Set `options(nanowasm.limits = wasm_limits(...))` to change the default for
every instance. Ctrl-C interrupts a running call, just as it does R code.

Every error nanowasm signals has class `nanowasm_error` and a more specific
class, so programs can react to each kind of failure:

```{r}
res <- tryCatch(
  slow$fib(40L),
  nanowasm_timeout = function(e) "too slow",
  nanowasm_trap = function(e) paste("trapped:", e$trap_id)
)
res
```

See `?"nanowasm-conditions"` for the full list: traps (with a `trap_id` such
as `"div_by_zero"` or `"unreachable"`), stack exhaustion, memory limits,
timeouts, link and validation errors, and errors in imported R functions.

## What the sandbox guarantees

A module can only affect the outside world through the imports you give it,
and nanowasm gives it none by default. It has no files, network, environment
variables, clock or random numbers unless an R function you import provides
them. A WASI environment provides some, and only those you ask for: the
arguments and environment variables you pass, the directories you grant
(read-only unless you say otherwise), R's clocks and random numbers, and
never sockets. What it can consume is bounded:

* **Memory**: everything the interpreter allocates for an instance
  (memories, tables and call stacks) counts against `wasm_limits(memory = )`.
* **Time**: each call is stopped after `wasm_limits(timeout = )`, and Ctrl-C
  always works.
* **Recursion**: deep calls end in a `nanowasm_stack_exhausted` error rather
  than a crash.

This makes nanowasm a sound way to run *untrusted computations*: a module
that loops, recurses or allocates without end is stopped with an error, and
R carries on. It is **not** a security boundary that has been audited
against deliberately hostile code. The interpreter and the glue around it
are tested, fuzzed and checked with sanitizers, but not formally verified.
Treat a module you don't trust as you would treat an R package you don't
trust, and don't give it imports that can do more than it needs.

## Getting modules

Any toolchain that targets WebAssembly can produce modules for nanowasm, as
long as the module needs no imports beyond WASI preview 1 and those you
provide:

* **The text format**, compiled with
  [wabt](https://github.com/WebAssembly/wabt)'s `wat2wasm`. This is how the
  example modules were made; their sources are installed next to them.
* **C**, with clang: `clang --target=wasm32 -nostdlib -O2
  -Wl,--no-entry -Wl,--export-all -o lib.wasm lib.c` for a library, or
  wasi-sdk's `clang --target=wasm32-wasip1 -O2 -o prog.wasm prog.c` for a
  program with the C standard library.
* **Rust**, with the `wasm32-unknown-unknown` target and a `cdylib` crate,
  or the `wasm32-wasip1` target for a program.
