---
title: "Examples"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A tour of what you can do with nanowasm, one small module at a time. Each
module is a few dozen lines of the WebAssembly text format, compiled with
`wat2wasm`; the sources are in
[`vignettes/articles/wasm/`](https://github.com/pedrobtz/nanowasm/tree/main/vignettes/articles/wasm)
in the package's repository. `vignette("nanowasm")` covers the API itself.

```{r setup}
library(nanowasm)

# Load one of this article's modules.
example_module <- function(name) {
  wasm_module(file.path("wasm", paste0(name, ".wasm")))
}
```

## Strings through linear memory

WebAssembly functions only take numbers, so text travels through the
module's memory: R writes the bytes, passes a pointer and a length, and reads
the result back. `strings.wasm` works on bytes in place:

```{r}
strings <- wasm_instantiate(example_module("strings"))
mem <- wasm_memory(strings)

text <- "hello from R, via WebAssembly"
len <- length(charToRaw(text))
wasm_write_string(mem, 0, text, nul = FALSE)

strings$count(0L, len, utf8ToInt("o"))
strings$upper(0L, len)
wasm_read_string(mem, 0, len)
```

Offset 0 is fine here because the module doesn't use its memory for
anything else. A real module usually exports an allocator; see the `sum.wasm`
example in `vignette("nanowasm")`.

## An image computed in WebAssembly

`mandelbrot.wasm` renders the Mandelbrot set into its memory, one byte per
pixel holding the number of iterations before the point escaped. R reads the
bytes back as a matrix and plots it:

```{r mandelbrot, fig.width = 7, fig.height = 5.25}
mandel <- wasm_instantiate(example_module("mandelbrot"))
w <- 240L
h <- 180L
maxiter <- 64L

mandel$render(0L, w, h, -2.4, -1.2, 0.9, 1.2, maxiter)
iterations <- wasm_read(wasm_memory(mandel), 0, w * h, "u8")
img <- matrix(iterations, nrow = w)

op <- par(mar = c(0, 0, 0, 0))
# Points that never escaped (the set itself) are black.
colours <- c(hcl.colors(maxiter, "Inferno"), "black")
image(img, col = colours, axes = FALSE, useRaster = TRUE)
par(op)
```

The module does all the arithmetic; R only handles the finished
`r w * h`-byte image.

## Calling R from WebAssembly

A module can import functions from its host. `integrate.wasm` implements the
trapezoidal rule for an imported `math.f`, so R decides what to integrate:

```{r}
wasm_imports(example_module("integrate"))

integrator <- function(f) {
  wasm_instantiate(
    example_module("integrate"),
    imports = list(math = list(f = wasm_func(f, params = "f64", results = "f64")))
  )
}

integrator(sin)$trapezoid(0, pi, 1000L)
integrator(dnorm)$trapezoid(-1.96, 1.96, 1000L)
```

The imported function can be any R closure, including one that keeps state.
Here it counts how often WebAssembly called back into R:

```{r}
calls <- 0
counted_exp <- function(x) {
  calls <<- calls + 1
  exp(-x^2)
}
integrator(counted_exp)$trapezoid(-3, 3, 200L)
calls
```

If the R function fails, the WebAssembly call stops and the error carries the
original condition:

```{r}
picky <- integrator(function(x) if (x > 1) stop("x is too large") else x)
err <- tryCatch(picky$trapezoid(0, 2, 10L), nanowasm_host_error = identity)
conditionMessage(err)
conditionMessage(err$parent)
```

## State lives in the instance

A module is code; an instance is that code with its own memory and globals.
Instances of the same module don't share state:

```{r}
counter <- example_module("counter")
a <- wasm_instantiate(counter)
b <- wasm_instantiate(counter)

for (i in 1:3) a$tick()
b$tick()

c(a = wasm_global(a, "count"), b = wasm_global(b, "count"))

wasm_global(b, "count") <- 100L
b$tick()
```

## Running code you don't trust

`untrusted.wasm` loops forever, recurses without end and grabs all the
memory it can. Under `wasm_limits()` each of those ends in an error of its
own class, and R carries on:

```{r}
untrusted <- wasm_instantiate(
  example_module("untrusted"),
  limits = wasm_limits(memory = 16 * 2^20, frames = 1000, timeout = 0.25)
)

run_safely <- function(expr) {
  tryCatch(
    expr,
    nanowasm_timeout = function(e) "stopped: time limit",
    nanowasm_stack_exhausted = function(e) "stopped: stack limit",
    nanowasm_memory_limit = function(e) "stopped: memory limit",
    nanowasm_trap = function(e) paste("trapped:", e$trap_id)
  )
}

run_safely(untrusted$spin())
run_safely(untrusted$recurse(0L))
```

A module asking for more memory than the limit allows is told no, as
WebAssembly specifies (`memory.grow` returns -1), so `hog()` stops growing
at the limit and returns the number of pages it managed to get:

```{r}
pages <- untrusted$hog()
pages
pages * 65536 / 2^20 # MiB
```

The instance is still usable after all of this:

```{r}
run_safely(untrusted$spin())
```
