---
title: "Real-world software"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The other articles use small modules written for the occasion. This one
runs real software, compiled unchanged to WebAssembly, from R:

* **SQLite**, the database engine, through its own command-line shell;
* **QuickJS**, a complete JavaScript engine;
* **md4c**, a Markdown-to-HTML renderer;
* a **C library** of terrain functions, called on an R matrix.

All of them run inside nanowasm's sandbox. The programs use WASI for their
input, output and files, and see only what each call grants them. The
binaries, their versions and licences are listed in
[`vignettes/articles/wasm/`](https://github.com/pedrobtz/nanowasm/tree/main/vignettes/articles/wasm)
in the package's repository.

```{r setup}
library(nanowasm)

article_module <- function(name) {
  wasm_module(file.path("wasm", paste0(name, ".wasm")))
}
```

## SQL on a CSV file with SQLite

`sqlite3.wasm` is the official SQLite shell (version 3.53.4), compiled from
the SQLite source with [wasi-sdk](https://github.com/WebAssembly/wasi-sdk).
Give it a directory and some SQL on standard input, and it can query the
files in that directory:

```{r}
sqlite_module <- article_module("sqlite3")

sqlite <- function(sql, dir, db = ":memory:", writable = FALSE) {
  # In a read-only directory SQLite can't create the lock file it uses on
  # WASI, so open database files as immutable there.
  if (!writable && db != ":memory:") db <- paste0("file:", db, "?immutable=1")
  wasm_run(
    sqlite_module,
    args = c("-csv", "-header", db),
    stdin = sql,
    dirs = dir,
    writable = writable,
    env = c(HOME = "/nowhere") # no ~/.sqliterc to look for
  )
}

data_dir <- file.path(tempdir(), "cars")
dir.create(data_dir, showWarnings = FALSE)
cars <- cbind(model = rownames(mtcars), mtcars)
write.csv(cars, file.path(data_dir, "mtcars.csv"), row.names = FALSE)

res <- sqlite(c(
  ".import mtcars.csv cars",
  "SELECT cyl, count(*) AS n, round(avg(mpg), 1) AS mpg, max(hp) AS max_hp",
  "FROM cars GROUP BY cyl ORDER BY cyl;"
), dir = data_dir)
read.csv(text = res$stdout)
```

With `writable = TRUE` it can create a database file, which stays on disk
for later runs:

```{r}
res <- sqlite(c(
  ".import mtcars.csv cars",
  "CREATE INDEX cars_by_mpg ON cars(mpg);"
), dir = data_dir, db = "cars.db", writable = TRUE)
res$status
list.files(data_dir)

# A later, read-only run queries it.
res <- sqlite(
  "SELECT model, mpg FROM cars ORDER BY mpg DESC LIMIT 3;",
  dir = data_dir, db = "cars.db"
)
read.csv(text = res$stdout)
```

Without `writable = TRUE` the program can't change anything, and it can
never see a file outside the directory it was given:

```{r}
sqlite("DELETE FROM cars;", dir = data_dir, db = "cars.db")$stderr
sqlite(".import ../../etc/passwd pw", dir = data_dir)$stderr
```

## JavaScript with QuickJS

`qjs.wasm` is [QuickJS-ng](https://github.com/quickjs-ng/quickjs)'s own
WASI build of its JavaScript engine, downloaded unmodified from its
release. A small helper turns it into a JavaScript evaluator:

```{r}
qjs_module <- article_module("qjs")

js <- function(code) {
  res <- wasm_run(qjs_module, args = c("-e", code))
  if (res$status != 0) stop(res$stderr, call. = FALSE)
  res$stdout
}

cat(js("
  const words = 'the quick brown fox jumps over the lazy dog'.split(' ');
  const byLength = Object.groupBy(words, w => w.length);
  console.log(JSON.stringify(byLength));
  console.log([...new Set(words)].sort().join(', '));
"))
```

Data goes in as JSON-ish text and comes back as text:

```{r}
x <- round(rnorm(8), 2)
code <- sprintf(
  "const x = [%s];
   const mean = x.reduce((a, b) => a + b) / x.length;
   console.log(mean.toFixed(4));",
  paste(x, collapse = ", ")
)
as.numeric(js(code))
round(mean(x), 4)
```

A JavaScript exception becomes an R error:

```{r, error = TRUE}
js("null.property")
```

## Markdown to HTML with md4c

`md2html.wasm` is [md4c](https://github.com/mity/md4c)'s command-line
renderer, which reads Markdown on standard input and writes HTML:

```{r}
markdown <- c(
  "### Rendered in WebAssembly",
  "",
  "This paragraph has *emphasis*, `code` and a [link](https://webassembly.org).",
  "",
  "| engine | language |",
  "|--------|----------|",
  "| SQLite | C        |",
  "| QuickJS| C        |"
)
html <- wasm_run(article_module("md2html"), args = "--ftables", stdin = markdown)$stdout
cat(html)
```

And here it is, rendered on this page:

```{r, echo = FALSE}
knitr::asis_output(html)
```

## A C library on an R matrix

Not every module is a program. `terrain.wasm` is a small C library, built
as a WASI "reactor": it has no `main`, but exports functions, plus its own
`alloc()` and `release()` for R to ask for space in its memory. Its
`hillshade()` computes how a landscape would look lit from the north-west,
the classic way to draw relief. `volcano`'s cells are 10 m wide; passing 5
exaggerates the relief twice, which makes it easier to see:

```{r}
terrain <- wasm_instantiate(article_module("terrain"), wasi = wasm_wasi())
wasm_exports(terrain)

# Run one of the library's grid functions on matrix z.
grid_fn <- function(fun, z, ...) {
  mem <- wasm_memory(terrain)
  n <- length(z)
  input <- terrain$alloc(8L * n)
  output <- terrain$alloc(8L * n)
  on.exit({
    terrain$release(input)
    terrain$release(output)
  })
  wasm_write(mem, input, as.vector(z), "f64")
  wasm_call(terrain, fun, input, output, nrow(z), ncol(z), ...)
  matrix(wasm_read(mem, output, n, "f64"), nrow = nrow(z))
}

shade <- grid_fn("hillshade", volcano, 5, 315, 45)
slope <- grid_fn("slope", volcano, 10)
summary(as.vector(slope))
```

```{r volcano, fig.width = 7, fig.height = 3.5}
op <- par(mfrow = c(1, 2), mar = c(0.5, 0.5, 2, 0.5))
image(volcano, col = terrain.colors(50), axes = FALSE, main = "Elevation")
image(shade, col = grey.colors(50, 0, 1), axes = FALSE, main = "Hillshade")
contour(volcano, add = TRUE, drawlabels = FALSE, col = "#00000040")
par(op)
```

## How fast is it?

nanowasm interprets WebAssembly rather than compiling it to machine code,
so it is slower than a native build or a compiling runtime. For programs
like these that is rarely what you notice:

```{r}
bench <- function(expr) round(system.time(expr)[["elapsed"]], 2)
c(
  sqlite_query = bench(sqlite("SELECT 1;", dir = data_dir)),
  javascript = bench(js("console.log(1)")),
  markdown = bench(wasm_run(article_module("md2html"), stdin = markdown)),
  hillshade = bench(grid_fn("hillshade", volcano, 5, 315, 45))
)
```

Times are in seconds, and include loading each module: SQLite and QuickJS
are about 1.5 MB of WebAssembly each.
